---
title: "R Review"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

For this review, we'll be working with dispatch data from the Lincoln Police Department. You can find the data online here: [http://opendata.lincoln.ne.gov/datasets/lpd-dispatch-records](http://opendata.lincoln.ne.gov/datasets/lpd-dispatch-records).

Use the Download menu to select a csv download.

Then move the data file you downloaded into your project folder.

### Load packages

Load the tidyverse, janitor and lubridate packages.

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(RColorBrewer)
```

### Load data

Load your data into an object called 'dispatches'.

```{r}
dispatches <- read_csv("LPD_Dispatch_Records.csv")
```

#### Fix column names

Use janitor to make all column names comply with R preferred style - all lowercase and underscores between words.

```{r}
dispatches %>% 
  clean_names() -> dispatches
```


### Analysis questions

#### Datatypes

Look at the documentation for the data. Do all the columns appear to be formatted correctly?
- No, date needs to be a date file and time should not be a character file it should probably be in hour:min format.

Use lubridate to change the RPT_Date column to ymd format.

```{r}
dispatches %>% 
  mutate(rpt_date_clean = ymd(rpt_date)) -> dispatches
```

Find the 13 rows that failed to parse. Why?
- In these rows all of the data is missing except an object id, and area of "Northeast Team" and a cfs_legend of other. The date has been listed as zero and  there is no case number. 

How many cases are in the data? How many unique cases?
- 400795 cases, 400774

```{r}
#failed to parse
dispatches %>% 
  filter(is.na(rpt_date_clean))

#total cases
nrow(dispatches)

#unique cases
length(unique(dispatches$case))
```

#### Arranging

What are the oldest and most recent cases in the data?
- oldest = case B7-000001, 01-01-2017, 00:00
- newest = case C0-051053, 06-07-2020, 23:59

```{r}
dispatches %>% 
  arrange(rpt_date_clean, rpt_time)

dispatches %>% 
  arrange(desc(rpt_date_clean, rpt_time))
```

#### Filtering

Create a dataframe called 'missing' with just missing persons cases in it.

```{r}
missing <- dispatches %>% 
  filter(cfs_legend == "MISSING PERSON")
```

Use the str_detect function to find all the cases that mention O Street in the address.

```{r}
dispatches %>% 
  filter(str_detect(block_address, "O ST"))
```

#### Counting

Use the count() function to find the number of dispatches to each neighborhood. 

```{r}
dispatches %>% 
  count(neighbor)%>% 
  arrange(desc(n))

```

Which neighborhood appears the most in the data?
- Downtown

Do you see any limitations of this data when we are counting by neighborhood?
- Counting by neighborhood does not account for the number of people, land area, types of businesses, ect. that many account for a large percentage of the variation in the number of dispatches by neighborhoods.

#### Mutating

Create a new column called 'year' that includes just the year from date_fixed.

```{r}
dispatches %>% 
  mutate(year = year(rpt_date_clean)) -> dispatches
```

Use the case_when function to create a new categorical variable dividing the rpt_time column into the 24 hours of the day. Be careful to make sure that each time fits into only one value of the new variable.

```{r}
dispatches %>% 
  mutate(rpt_time = as.numeric(rpt_time)) %>% 
  mutate(hour = case_when(
    rpt_time <  0100 ~ 0,
    rpt_time >= 0100 & rpt_time <0200 ~ 1,
    rpt_time >= 0200 & rpt_time <0300 ~ 2,
    rpt_time >= 0300 & rpt_time <0400 ~ 3,
    rpt_time >= 0400 & rpt_time <0500 ~ 4,
    rpt_time >= 0500 & rpt_time <0600 ~ 5,
    rpt_time >= 0600 & rpt_time <0700 ~ 6,
    rpt_time >= 0700 & rpt_time <0800 ~ 7,
    rpt_time >= 0900 & rpt_time <1000 ~ 9,
    rpt_time >= 1000 & rpt_time <1100 ~ 10,
    rpt_time >= 1100 & rpt_time <1200 ~ 11,
    rpt_time >= 1200 & rpt_time <1300 ~ 12,
    rpt_time >= 1300 & rpt_time <1400 ~ 13,
    rpt_time >= 1400 & rpt_time <1500 ~ 14,
    rpt_time >= 1500 & rpt_time <1600 ~ 15,
    rpt_time >= 1600 & rpt_time <1700 ~ 16,
    rpt_time >= 1700 & rpt_time <1800 ~ 17,
    rpt_time >= 1800 & rpt_time <1900 ~ 18,
    rpt_time >= 1900 & rpt_time <2000 ~ 19,
    rpt_time >= 2000 & rpt_time <2100 ~ 20,
    rpt_time >= 2100 & rpt_time <2200 ~ 21,
    rpt_time >= 2200 & rpt_time <2300 ~ 22,
    rpt_time >= 2300 & rpt_time <2400 ~ 23,
    rpt_time >= 2400 & rpt_time <2500 ~ 24,
    )) %>% 
  count(hour) %>% 
  arrange(desc(n))


#alternative method
dispatches %>% 
  separate(rpt_time, into =c('hour', 'min'), sep=2, remove =F) %>% 
  count(hour) %>% 
  arrange(desc(n))

```

What hour of the day sees the most police action?
16 or 4PM

#### Grouping and summarizing 

How many drug-related dispatches occurred in each year of the data?

```{r}
dispatches %>% 
  filter(cfs_legend=="DRUGS") %>% 
  count(year)
  
```

Create a new column called month. Then using group_by and summarize, find the maximum, minimum and average number of dispatches per month.

```{r}
dispatches %>%
  mutate(month = month(rpt_date_clean, label=T)) -> dispatches

dispatches%>% 
  group_by(year, month) %>% 
  count(month) %>% 
  group_by(month) %>% 
  summarize(average = mean(n), min = min(n), max=max(n))
```


#### Percent change

What was the percent change in total number of dispatches from 2018 to 2019?
0.827%

```{r}
dispatches %>% 
  count(year)

((118088-117120)/117120)*100
  
```


#### Line charts

Using ggplot, create a line chart of the number of cases per month and year. Choose an appropriate color for the line, add a title and labels and a choose a theme.

```{r}

dispatches %>% 
  filter(month != "NA") %>% 
  filter(rpt_date_clean < ymd("2020/05/30")) %>% 
  group_by(year, month) %>% 
  count(month) %>%  
  ggplot(aes(x=month, y=n, group=factor(year), color=factor(year)))+
           geom_line()+
  geom_point()+
  scale_color_brewer(palette="Oranges")+
  labs(x="", y="",color="Year", title = "Police Dispatch Calls by Month in Lincoln, Nebraska")+
  theme(legend.title=element_blank())+
  theme_classic()



```

What do you observe about the yearly pattern of police dispatches? Why do you suppose that is?

#### Column charts

Using ggplot, create a column chart that shows the five most common categories in the cfs_legend column. Apply appropriate design decisions to your chart.

```{r}
dispatches %>% 
  group_by(cfs_legend) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n)) %>% 
  top_n(5) ->top5

order <- c("OTHER", "DISTURBANCE", "TRAFFIC", "ACCIDENT", "SUSPICIOUS") 
  
top5 %>% ggplot(aes(x=cfs_legend, y=n))+
  scale_x_discrete(limits=order, labels=c("Other", "Disturbance", "Traffic", "Accident", "Suspicious"))+
  geom_col(fill="skyblue3")+
  labs(x="", y="", title="Five Most Common Reasons for Police Dispatch in Lincoln, Nebraska")+
  theme_classic()
  

```

